---
title: "Sampling from User-Defined Distributions using stors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sampling_from_User_Defined_Distributions_using_stors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `stors` package is designed to generate samples from uni- and multimodal distributions using Rejection Sampling techniques. While it provides optimized sampling functions for built-in distributions, STORS also offers flexibility for users to define their own distributions and create custom proposals for efficient sampling.

In this vignette, we focus on how to use the `stors` package to sample from user-defined distributions. We will demonstrate how to build custom proposals for distributions not covered by the built-in functions and how to utilize the package's functions to sample efficiently from these distributions.

To sample from user-defined distributions, you can create custom proposals using the `build_proposal()` function and then use the `build_proposal()` function to create a sampler based on the custom proposal.


## Sampling from User-Defined Distributions

For distributions not covered by the built-in functions and that satisfy the log-concavity condition, users have the flexibility to create their own proposals using the `build_proposal()` function. This function is designed to handle a variety of distributions by accepting a range of parameters tailored to the specific needs of the distribution being sampled from.

The following example demonstrates the process of sampling from a bimodal normal distribution. Let's assume that you want to sample from a bimodal distribution, which is a combination of two normal distributions \( f(x) = 0.5 w_1(x) + 0.5 w_2(x) \), where \( w_1(x) \sim N(0, 1) \) and \( w_2(x) \sim N(4, 1) \):

\[
f(x) =  0.5 \frac{1}{\sqrt{2 \pi}}  \exp\left(-\frac{x^2}{2}\right) + 0.5  \frac{1}{\sqrt{2 \pi}} \exp\left(-\frac{(x - 4)^2}{2}\right)
\]

To sample from the tails of \( f(x) \) using ARS, we need the log transformation of \( f(x) \), \( h(x) = \log(f(x)) \), and its first derivative \( h_{\text{prime}}(x) = \frac{d}{dx} h(x) \). Also, the modes of \( f(x) \) are required to build the steps part of the proposal distribution, which in this case are approximately at \( x = 0 \) and \( x = 4 \).


```{r , echo = TRUE}

library(stors)

modes_bi <- c(0.00134865, 3.99865)

f_bi <- function(x) {
  0.5 * (sqrt(2 * pi))^(-1) * exp(-(x^2) / 2) + 0.5 * (sqrt(2 * pi))^(-1) * exp(-((x - 4)^2) / 2)
}

h_bi <- function(x) log(f_bi(x))


h_prime_bi <- function(x) {
  (-(exp(-1 / 2 * (-4 + x)^2) * 0.5 * (-4 + x)) / sqrt(2 * pi) - (exp(-x^2 / 2) * 0.5 * x) / sqrt(2 *  pi)) / ((exp(-x^2 / 2) * 0.5) / sqrt(2 * pi) + (exp(-1 / 2 * (-4 + x)^2) * 0.5) / sqrt(2 * pi))
}

bi_proposal <- build_proposal(lower = -Inf, upper = Inf, modes = modes_bi, f = f_bi, h = h_bi, h_prime = h_prime_bi)

plot(bi_proposal)

```

After building the desired proposal with `build_proposal()`, users can explore its properties using the `print()` function and visualize its structure using the `plot()` function. This step is similar to what was done previously with the built-in proposal and helps in understanding the proposal's configuration and ensuring that it aligns with the requirements of the target distribution.

```{r , echo = TRUE}

# Printing the properties of the custom proposal
print(bi_proposal)

# Visualizing the custom proposal
plot(bi_proposal)

```

Once the proposal is configured and validated, it can be passed to the `build_sampler()` function. `build_sampler()` caches the proposal in memory, providing fast access for sampling, and returns a sampler function for the target distribution. This sampler function can then be used to generate samples as needed.

```{r , echo = TRUE}

# Creating a sampler function using the custom proposal
bi_sampler <- build_sampler(bi_proposal)

plot(bi_proposal)

# Generating samples from the target distribution
sample <- bi_sampler(10^3)

hist(sample, main = "Sample Generated From Multi-modal Distribution ", xlab = "Value", col = "lightblue", border = "white")


```

Sampling from a Truncated User-Defined Distribution
To sample from a truncated user-defined distribution, simply provide the lower and upper truncation bounds using the \code{lower} (lower bound) and \code{upper} (upper bound) arguments. Below, we demonstrate this process using a previously created multi-modal distribution.

Example: Sampling from a Truncated Multi-Modal Distribution
```{r , echo = TRUE}

# Define modes for the multi-modal distribution
modes_bi <- c(0.00134865, 3.99865)

# Build a truncated proposal for the multi-modal distribution
bi_trunc_proposal <- build_proposal(
  lower = -1,
  upper = 6,
  modes = modes_bi,
  f = f_bi,
  h = h_bi,
  h_prime = h_prime_bi,
  steps = 2040
)

# Visualize the truncated proposal
plot(bi_trunc_proposal, main = "Truncated Multi-Modal Proposal (-1, 6)", xlab = "Value", col = "skyblue")

# Create a sampler for the truncated distribution
bi_sampler_trunc_sampler <- build_sampler(bi_trunc_proposal)

# Generate 10^3 samples from the truncated distribution
sample <- bi_sampler_trunc_sampler(10^3)

# Generate 10^3 samples from the truncated distribution
sample <- bi_sampler_trunc_sampler(10^3)

# Visualize the generated samples
hist(
  sample,
  main = "Histogram of Samples from Truncated Multi-Modal Distribution",
  xlab = "Value",
  col = "lightgreen",
  border = "white"
)



```


## Saving, Loading and Deleting Users' Proposals

The proposal stored in `bi_proposal` will not be preserved across sessions. Therefore, if the user is satisfied with the sampling results and wishes to store the proposal for future use, they should save the proposal using the `save_proposal()` function. This function takes the optimized proposal and a filename chosen by the user to save the proposal in the internal R data directory on their machine. This is recommended if the user plans to use the proposal frequently across sessions, to avoid re-optimizing the proposal in future sessions.

```{r , echo = TRUE}

save_proposal(bi_proposal, "bimodal_proposal")

```

```{r , include = FALSE}
for (i in 1:3) {
  save_proposal(bi_proposal, paste0("bimodal_proposal_", i))
}

```

To load and use the proposal in future sessions, the user only needs to use the `load_proposal()` function.

```{r , echo = TRUE}

bi_proposal <- load_proposal("bimodal_proposal")

```

In case the user wants to view a list of all previously stored proposals, the `print_proposals()` function will provide the user with all stored proposals.

```{r , echo = TRUE}

print_proposals()

```

Be aware that the saved proposal can be quite large in size, depending on the user's system. Therefore, if the user wants to delete a proposal, the `delete_proposal()` function can be used by providing the proposal's name.

```{r , echo = TRUE}

delete_proposal("bimodal_proposal")

print_proposals()

```

```{r , include = FALSE}
for (i in 1:3) {
  delete_proposal(paste0("bimodal_proposal_", i))
}
```
