---
title: "Sampling from Built-in Distributions using STORS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sampling_Built_in_distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `stors` package provides efficient sampling from uni- and multimodal distributions using Rejection Sampling techniques. For built-in distributions, STORS offers optimized sampling functions that leverage pre-constructed grids to maximize sampling speed.

In this vignette, we focus on how to use the `stors` package to sample from built-in distributions, demonstrating the simplicity and efficiency of the package's built-in functions.

To use the `stors` package with built-in distributions, you only need to run the `grid_optimizer()` function once when the package is loaded for the first time. This creates an optimized grid and stores it in the internal R data directory, so it can be loaded and used for sampling when needed. You can also visualize the grid data using the `plot()` function.

Here's an example of how to optimize a grid for the standard normal distribution:

```{r setup, echo = TRUE}
library(stors)

norm_grid = srnorm_optimize()
```

With the optimized grid ready, sampling from the standard normal distribution can be easily achieved using the `srnorm` function. For general normal distributions, where the mean and standard deviation are not zero and one respectively, the `srnorm_scaled` function can be used.

```{r , echo = TRUE}
# Generating samples from the standard normal distribution
standard_normal_samples = srnorm(10)
print(standard_normal_samples)

# Generating samples from a normal distribution with mean = 4 and standard deviation = 2
# general_normal_samples = srnorm_scaled(10, mu = 4, sd = 2)
# print(general_normal_samples)
```

One of the powerful features of the `stors` package is its ability to sample from truncated distributions efficiently. This is particularly useful when the area of interest in a distribution is limited to a specific range, and you want to avoid unnecessary computations outside this range.

The `stors` package simplifies the process of sampling from truncated distributions. For standard normal distributions, the `srnorm_truncate()` function enables this feature. Users just need to provide the truncation bounds, and the function returns a sampler that is confined to these bounds. This means that the sampling is focused only within the specified range, making the process more efficient, especially when the cumulative distribution function (CDF) does not need to be evaluated.

Here's an example of how to use `srnorm_truncate()` to create a sampler for a truncated normal distribution:

```{r , echo = TRUE}
# Defining the truncation bounds
lower_bound = -1
upper_bound = 1

# Creating a sampling function for a truncated normal distribution
truncated_normal_sampler = srnorm_truncate(xl = lower_bound, xr = upper_bound)

# Generating samples from the truncated normal distribution
truncated_samples = truncated_normal_sampler(1000)
hist(truncated_samples)
```

The `grid_optimizer()` function not only optimizes the grid but also returns a list detailing the properties of the optimized grid. This list can be used to understand the grid's structure and properties. To view these details, you can use the base `print()` function. Additionally, the `plot()` function provides a visual representation of the grid.

```{r , echo = TRUE}
print(norm_grid)

plot(norm_grid)
```

In cases where the grid is very dense and challenging to visualize in its entirety, you can focus on a specific region of the grid. The `plot()` function allows you to specify the limits on the x-axis using `x_min` and `x_max` parameters. This can be particularly useful for zooming into areas of interest within the grid.

```{r , echo = TRUE}
plot(norm_grid, x_min = -0.1, x_max = 0)
```

These functionalities of the `grid_optimizer()` and the subsequent visualization tools provide users with in-depth insights into the characteristics of the generated grid, enabling them to understand and fine-tune the sampling process.

If you are not satisfied with the automatically optimized grid, the `grid_optimizer()` function offers several arguments for customization. One such argument is `target_sample_size`, which defaults to 1000. If you wish to optimize the grid for significantly larger or smaller sample sizes, this argument can be altered.

Another important argument is the pre-acceptance threshold `theta`, which essentially is the ratio between the lower and upper linear bounds for each step. This is typically set to a very small number close to zero. The reason is that the cost of rejecting samples, even in heavy-tailed distributions, is usually less than the cost of using alternative exact sampling methods (like Inverse Transform) for dense grids, as the probability of hitting these inefficient steps is low. However, for very large sample sizes (e.g., \(10^9\)), these inefficient steps might be hit more frequently, leading to an increase in sampling time due to lower efficiency. In such cases, you can set a higher pre-acceptance threshold using the `theta` parameter.

For users who are less familiar with the target distribution or want to focus on a specific range of the distribution, the `grid_range` argument can be used. This parameter limits the step construction to a specified range on the x-axis, offering more control over the grid structure.

```{r , echo = TRUE}
# Customizing the grid for a specific sample size
custom_grid = srnorm_optimize( target_sample_size = 10000)

print(custom_grid)

plot(custom_grid)

# Customizing the grid with a specific pre-acceptance threshold
custom_grid_high_theta = srnorm_optimize( theta = 0.9)

print(custom_grid_high_theta)

plot(custom_grid_high_theta)

# Customizing the grid within a specific range
custom_grid_range = srnorm_optimize( grid_range = c(-1, 1))

print(custom_grid_range)

plot(custom_grid_range)
```

In addition to the `target_sample_size`, `theta`, and `grid_range` parameters, users have the flexibility to directly control the number of steps in the proposal grid. This can be particularly useful when you have specific requirements or preferences for the granularity of the grid. To specify the number of steps, the `steps` argument can be passed to the `grid_optimizer()` function.

By setting the `steps` parameter, you can determine the resolution of the grid, which can impact the balance between computational efficiency and the granularity of the sampling. A higher number of steps typically provides a more detailed grid, potentially leading to more precise sampling, but at the cost of increased memory usage and possibly longer computation times.

```{r , echo = TRUE}
# Customizing the grid with a specific number of steps
custom_grid_steps = srnorm_optimize( steps = 50)

plot(custom_grid_steps)

custom_grid_steps = srnorm_optimize( steps = 500)

plot(custom_grid_steps)
```

There are many built-in functions in `stors`; the list below shows all available functions that can be used directly by the user after optimizing the grids:

```{r, echo = TRUE}
# List of available built-in sampling functions
print(stors:::stors_env$grids$builtin$names)
```

